A new type of automated prover based on category theory

### 2022-08-11 paper.tex
revision 1

\usepackage{amsmath, amssymb, graphics, setspace}
\usepackage[letterpaper, total={6.5in, 9in}]{geometry}
\usepackage{cite}
\usepackage{tikz}
\usepackage{tikz-cd}
\usepackage{float, multicol}
\usetikzlibrary{positioning, cd}

\newcommand{\mathsym}[1]{{}}
\newcommand{\unicode}[1]{{}}
\newtheorem{theorem}{Theorem}
\newtheorem{axiom}{Axiom}
\newtheorem{definition}{Definition}
%% \newtheorem{example}{Example}

\setlength{\parskip}{0.8em}

@@

\title{A new type of automated prover based on category theory}
\author{Zijian Wang}
\date{}
\maketitle

@@

\hrule
\begin{abstract}
In this article I present a new type of automated statement proof algorithm based on new data structures, i.e. brackets and map graphs
and new algorithms. The brackets provide an elegant low-knowledge representation of mathematical concepts. The map graphs offer an efficient machine-learning method which let the computer learn knowledge while proving. Additionally, the new finding is totally built on the category theory. Furthermore, a prototype of the program is given.
\end{abstract}
\hrule

\section*{Introduction}
During the long period of development of the research on automated theorem provers, there have already existed a great number of research and implementation on proofs focusing on a specific mathematical subject, such as algebra equations (for example, Wolfram Mathematica{'}s FindEquationProof function \cite{2}) or geometrical theorems (for example, the Geometry Expert or GEX \cite{3} of Key Laboratory of Mathematics Mechanization, Chinese Academy of Sciences). Some relatively new works in this aspect includes Microsoft{'}s Lean prover and Z3 algorithm.

@@

\section*{Preliminaries}

\subsection*{Notations and Conventions}

@@

where \(A\) and \(B\) are sets and\(x\)and \(y\) are elements in \(A\) and \(B\).

@@

\subsection*{Pre-Knowledge}

Readers should have basic understandings of the category theory \cite{1} and axiomatic set theory. At least, they should know the basic concepts as well as these conclusions and theorems:

@@

	\pmb{ Proof:} First, we select element \(\mho \notin S\). As the ZFC system{'}s selection axiom enables to select element \(g(S')\) from each subset \(S'\) of \(S\). Let \(a_0\text{:=}g(S)\). Using transfinite recursive theorem to each ordinal \(\alpha\) we define:

	\[a_{\alpha }\text{:=}\left\{
	\begin{array}{cc}
		g\left(S\backslash \left\{\left.a_{\beta }\right|\beta <\alpha \right\}\right), & S\neq \left\{\left.a_{\beta }\right|\beta <\alpha \right\} \\
		\mho , & S=\left\{\left.a_{\beta }\right|\beta <\alpha \right\} \\
	\end{array}
	\right.\]

	Then, since \(S\) is a set, a minimized \(\theta\) can be chosen making \(a_{\theta }=\mho , S=\left\{\left.a_{\beta }\right|\beta <\theta \right\}\) and significantly there exists a bijection between \(S\) and \(\theta =\{\beta |\beta <\theta \}\). Following the definition, there obviously exists a well-ordering on \(S\).
	\begin{flushright}
		$\blacksquare $
	\end{flushright}
\end{theorem}

@@

\[B=\frac{100 k}{(\ln  t)^c+k}\]

Where \(B\) is percent of memorization while\(t\)is time and \(c=1.25,k=1.84\). This curve is used by the algorithm for self-optimization purposes.

@@

\section*{The Computerized Representation of Mathematical Structures}

Before more specific discussion, we select Grothendieck universe \(\mathcal{U}\) \cite{1} to avoid set theory paradoxes. Besides, we define meta category \(\mathcal{C}\) where all discussions take place. There are briefly three kind of objects in \(\mathcal{C}\): symbols, notations and brackets. All symbols in \(\mathcal{C}\) forms sub category \(\text{Sym}(\mathcal{C})\) called the \textbf{ Symbol Subcategory}. All notations in \(\mathcal{C}\) forms \(\text{Not}(\mathcal{C})\) and all brackets forms \(\text{Bra}(\mathcal{C})\), called the \textbf{ Notation Subcategory} and the \textbf{ Bracket Subcategory}. That is,

@@

%% \pmb{ Definition.} 
\begin{definition}
	A \textbf{ notation} \(\mathcal{N}\) is defined as an object in \(\text{Not}(\mathcal{C})\) which is a sub category of \(\mathcal{C}\). The \(\text{Ob}(\text{Not}(\mathcal{C}))\) set can be mapped to \(\text{On}_{\geq 0}\) class, indicating the number of parameters that a notation take. In which,

@@

%% \pmb{ Definition.}
\begin{definition}
	A \textbf{ symbol} \(\mathcal{S}\) is defined as an object in \(\text{Sym}(\mathcal{C})\) which is a sub category of \(\mathcal{C}\). There exists a functor from \(\text{Sym}(\mathcal{C})\) to \(\text{Not}(\mathcal{C})\):

@@

	Where \(\nu (\mathcal{S})\) is the \textbf{ corresponding notation} of \(\mathcal{S}\) and must satisfy \(\forall \mathcal{S}\in \text{Sym}(\mathcal{C}),\mathit{p}(\nu (\mathcal{S}))=0\), indicating all corresponding notations are types instead of operators. For convenience and precision, the text that meaning creating a symbol \(\mathcal{S}\) whose corresponding notation is \(\mathcal{N}\)can and should be written in the form below.

@@

\begin{table}
	\centering
	\[\begin{array}{|c|c|c|c|c|c|c|c|c|c|c|}
		\hline
		\mathcal{N} & \text{integer} & \text{number} & \text{infinity} & \sum  & = & / & - & \text{abs} & < & \text{for} \\
		\hline
		\mathit{p}(\mathcal{N}) & 0 & 0 & 0 & 3 & 2 & 2 & 2 & 1 & 2 & 2 \\
		\hline
	\end{array}\]
	\caption{The table for $\mathit{p}(\mathcal{N})$}
\end{table}

@@

	It is essential to pay attention to that a bracket with 2-parameter notation is not necessarily a micro-statement. This conclusion can be illustrated via the example of \(\{(0,+),(1,x),(2,y)\}\) whose notation has two symbols as parameters but is not a micro-statement. The second restriction is more important, which is that a micro-statement must have the capacity to be evaluated into a Boolean value, that is, \(\text{true}\) or \(\text{false}\). What is notable, the \textbf{ evaluation} process is not declared inside the representation system, whose definition comes below. All micro-statement brackets forms the \textbf{ Micro-Statement }sub category of \(\text{Bra}(\mathcal{C})\) and is annotated as \(\text{Mic}(\mathcal{C})\).
\end{definition}

@@

\section*{The Standardized Principles for Derivation of Statements}

@@

	Where \(\mathcal{M}_i\) { }is a micro-statement pool and for each element in \(\mathcal{D}\), its left pair element is a unique ordinal. The $\mathcal{D}$ set is well-ordered following the sort of \(i\). On the beginning, the the initial pool equals to the condition, that is the micro-statements given by the problem. On the other hand, the conclusions of the proof problem, to be proved via some steps form a set \(\mathcal{R}\), called the \textbf{requirement set} are to be inferred to exist inside the micro-statement pool.

@@

Then each \(\mathit{r}\) must belongs to region \(\mathbb{Z}_{\left.\left[0,2^{32}\right.\right)}\). That is, obviously, if \(\sum _{u=0}^N \left|\text{Hom}_{\mathcal{Q}}\left(\mathcal{K}_j,\mathcal{K}_q\right)\right|\geq 2^{32}\), then there will be some of the functors that will never be referred to, although there do exist morphisms targeting them. While in this paper the problem will not be solved and we assume that the number of the morphisms added up is smaller than the boundary of \(2^{32}\), which equals to four giga bytes and will not be exceeded in most engineering situations. Obviously although using the portable recursion equation the quality of the random numbers generated may be lower than nowadays{'} some technical implementations, it can be conveniently expressed in this paper. If the algorithm is implemented into the read world, a randomness collector, which generates random numbers from the real world, for example, the viberation of the computer motherboard, will probably act as a better solution.

@@

A situation after some reinforcements may be the one demonstrated in the figure on the next page.

\begin{figure}[ht]
	\centering
	\begin{tikzpicture}[commutative diagrams/every diagram]
		\node (K0) at (90:2.3cm) {$\mathcal{K}_0$};
		\node (K1) at (90+72:2cm) {$\mathcal{K}_1$};
		\node (K2) at (90+2*72:2cm) {$\mathcal{K}_2$};
		\node (K3) at (90+3*72:2cm) {$\mathcal{K}_3$};
		\node (Kn) at (90+4*72:2cm) {$\mathcal{K}_N$};

@@

@@

\section*{Conclusions and Future Work}

@@

\section*{Acknowledgments}
I express my sincerest greatest gratitude to my tutor Prof. Shao Xinhui, the Yingcai Project of P.R.China, which enables senior high school students to conduct science research, my senior high academy Northeast Yucai School (NEYC) of P.R.China and the corresponding university Northern Eastern University (NEU) of P.R.China. I also thank my parents for encouraging me to conduct this research project.

\bibliographystyle{acm}
\bibliography{document}
\end{document}

>>>>

\usepackage{amsmath, amssymb, graphics, setspace}	%% for maths display
\usepackage[letterpaper, total={6.5in, 9in}]{geometry}	%% set paper size & margin
\usepackage{cite}	%% for citations
\usepackage{tikz, tikz-cd}	%% for commutative diagrams
\usepackage{flowchart}	%% for flowchart
\usepackage{float, booktabs}	%% for table & alignment
\usepackage{algorithmicx, algorithm, algpseudocode}	%% for pseudo code composition
\usetikzlibrary{positioning, cd, arrows}	%% for comutative diagram & flowchart drawing

\newcommand{\mathsym}[1]{{}}	%% generated by mma
\newcommand{\unicode}[1]{{}}	%% generated by mma
\newtheorem{theorem}{Theorem}	%% definition for theorem env
\newtheorem{axiom}{Axiom}	%% definition for axiom env
\newtheorem{definition}{Definition}	%% definition for definition env

\numberwithin{theorem}{section}	%% count theorems
\numberwithin{axiom}{section}	%% count axioms
\numberwithin{definition}{section}	%% count definitions

\setlength{\parskip}{0.8em}	%% set white space between paragraphs

@@

\title{A new type of automated prover based on category theory}
\author{Zijian Wang}
\date{\today}	%% compile date
\maketitle

@@

\hrule %% horizonal line
\begin{abstract}
In this article a new type of automated statement proof algorithm based on new data structures, i.e. brackets and map graphs
and new algorithms is presented. The brackets provide an elegant low-knowledge representation of mathematical concepts. The map graphs offer an efficient machine-learning method which let the computer learn knowledge while proving. Additionally, the new finding is totally built on the category theory. Furthermore, a prototype of the program is given.
\end{abstract}
\hrule	%% horizonal line

\section{Introduction}
During the long period of development of the research on automated theorem provers, there have already existed a great number of research and implementation on proofs focusing on a specific mathematical subject, such as algebra equations (for example, Wolfram Mathematica{'}s FindEquationProof function \cite{2}) or geometrical theorems (for example, the Geometry Expert or GEX \cite{3} of Key Laboratory of Mathematics Mechanization, Chinese Academy of Sciences). Some relatively new works in this aspect includes Microsoft{'}s Lean prover \cite{8} and Z3 \cite{9} algorithm.

@@

\section{Preliminaries}

\subsection{Notations and Conventions}

@@

\noindent where \(A\) and \(B\) are sets and\(x\)and \(y\) are elements in \(A\) and \(B\).

@@

\subsection{Pre-Knowledge}

The definitions and theories presented in the paper utilizes the concepts of the category theory \cite{1} and axiomatic set theory \cite{1}. Therefore, readers are recommended to review understandings and conclusions about the two concepts before going over the paper.

@@

	The proof can be found in citation \cite{1}.
\end{theorem}

@@

\[E(\mathit{t})=\frac{100 k}{(\ln  t)^c+k}\]

\noindent where \(B\) is percent of memorization while\(t\)is time and \(c=1.25,k=1.84\). This curve is used by the algorithm for self-optimization purposes.

@@

\section{The Computerized Representation of Mathematical Structures}

Before more specific discussion, we select Grothendieck universe \(\mathcal{U}\) \cite{1} to avoid set theory paradoxes. Besides, we define meta category \(\mathcal{C}\) where all discussions take place. There are briefly three kind of objects in \(\mathcal{C}\): symbols, notations and brackets. All symbols in \(\mathcal{C}\) forms sub-category \(\text{Sym}(\mathcal{C})\) called the \textbf{ Symbol Subcategory}. All notations in \(\mathcal{C}\) forms \(\text{Not}(\mathcal{C})\) and all brackets forms \(\text{Bra}(\mathcal{C})\), called the \textbf{ Notation Subcategory} and the \textbf{ Bracket Subcategory}. That is,

@@

%% \pmb{ Definition.} 
\begin{definition}
	A \textbf{ notation} \(\mathcal{N}\) is defined as an object in \(\text{Not}(\mathcal{C})\) which is a sub-category of \(\mathcal{C}\). The \(\text{Ob}(\text{Not}(\mathcal{C}))\) set can be mapped to \(\text{On}_{\geq 0}\) class, indicating the number of parameters that a notation take. In which,

@@

%% \pmb{ Definition.}
\begin{definition}
	A \textbf{ symbol} \(\mathcal{S}\) is defined as an object in \(\text{Sym}(\mathcal{C})\) which is a sub-category of \(\mathcal{C}\). There exists a functor from \(\text{Sym}(\mathcal{C})\) to \(\text{Not}(\mathcal{C})\):

@@

	\noindent where \(\nu (\mathcal{S})\) is the \textbf{ corresponding notation} of \(\mathcal{S}\) and must satisfy \(\forall \mathcal{S}\in \text{Sym}(\mathcal{C}),\mathit{p}(\nu (\mathcal{S}))=0\), indicating all corresponding notations are types instead of operators. For convenience and precision, the text that meaning creating a symbol \(\mathcal{S}\) whose corresponding notation is \(\mathcal{N}\)can and should be written in the form below.

@@

\begin{table}
	\centering
	\begin{tabular}{ccccccccccc}
		\toprule
		$\mathcal{N}$ & integer & number & infinity & $\sum$ & $=$ & $/$ & - & abs & $<$ & for \\
		\midrule
		$\mathit{p} (\mathcal{N})$ & $0$ & $0$ & $0$ & $3$ & $2$ & $2$ & $2$ & $1$ & $2$ & $2$ \\
		\bottomrule
	\end{tabular}
	\caption{The table for $\mathit{p}(\mathcal{N})$}
\end{table}

@@

	It is essential to pay attention to that a bracket with 2-parameter notation is not necessarily a micro-statement. This conclusion can be illustrated via the example of \(\{(0,+),(1,x),(2,y)\}\) whose notation has two symbols as parameters but is not a micro-statement. The second restriction is more important, which is that a micro-statement must have the capacity to be evaluated into a Boolean value, that is, \(\text{true}\) or \(\text{false}\). What is notable, the \textbf{ evaluation} process is not declared inside the representation system, whose definition comes below. All micro-statement brackets forms the \textbf{ Micro-Statement }sub-category of \(\text{Bra}(\mathcal{C})\) and is annotated as \(\text{Mic}(\mathcal{C})\).
\end{definition}

@@

\section{The Standardized Principles for Derivation of Statements}

@@

	\noindent where \(\mathcal{M}_i\) { }is a micro-statement pool and for each element in \(\mathcal{D}\), its left pair element is a unique ordinal. The $\mathcal{D}$ set is well-ordered following the sort of \(i\). On the beginning, the the initial pool equals to the condition, that is the micro-statements given by the problem. On the other hand, the conclusions of the proof problem, to be proved via some steps form a set \(\mathcal{R}\), called the \textbf{requirement set} are to be inferred to exist inside the micro-statement pool.

@@

Then each \(\mathit{r}\) must belongs to region \(\mathbb{Z}_{\left.\left[0,2^{32}\right.\right)}\). That is, obviously, if \(\sum _{u=0}^N \left|\text{Hom}_{\mathcal{Q}}\left(\mathcal{K}_j,\mathcal{K}_q\right)\right|\geq 2^{32}\), then there will be some of the functors that will never be referred to, although there do exist morphisms targeting them. While in this paper the problem will not be solved and we assume that the number of the morphisms added up is smaller than the boundary of \(2^{32}\), which equals to four giga bytes and will not be exceeded in most engineering situations. Obviously although using the portable recursion equation the quality of the random numbers generated may be lower than nowadays{'} some technical implementations, it can be conveniently expressed in this paper. If the algorithm is implemented into the read world, a randomness collector, which generates random numbers from the real world, for example, the vibration of the computer motherboard, will probably act as a better solution.

The more detailed explanation of the recursive functor can be found in the algorithm below.

\begin{algorithm}[h]
	\caption{Pseudo code for recursive functor $\mathcal{F}_t$}
	\begin{algorithmic}[1]
		\Require Pool $\mathcal{M}_{t}$, Knowledge base $\mathcal{Q}$, random $\mathit{r}$, Just used reflection functor index $i$
		\Ensure Next pool $\mathcal{M}_{t+1}$

		\Function{$\mathcal{F}_t$}{$\mathcal{M}_t$, $\mathcal{Q}$, $\mathit{r}$, $i$}
			\Comment{This is the sub-process}
			\State $\mathcal{K} \gets $ \Call{SelectReflectionFunctor}{$\mathcal{Q}$, $\mathit{r}$, $i$}
			%% \State \Return \Call{Reflect}{$\mathcal{K}$, $\mathcal{M}_{t}$}
			\State \Return $\mathcal{K} \mathcal{M}_t$
			\Comment{Apply functor $\mathcal{K}$ to pool}
		\EndFunction

		\Function{SelectReflectionFunctor}{$\mathcal{Q}$, $\mathit{r}$, $i$}
			\Comment{Just used: $\mathcal{K}_i \in \text{Ob(}\mathcal{Q}\text{)}$}
			\State $M \gets Mor(\mathcal{Q})$
			\ForAll{$m \in M$}
				\Comment{Get morphisms starting at $\mathcal{K}_i$}
				\If{$s(m) != \mathcal{K}_i$}
					\State $M \gets M \backslash \{m\}$
				\EndIf
			\EndFor
			\If{$|M|==1$}
				\Comment{The easiest condition}
				\State \Return $t(m \in M)$
			\Else
				\State $n \gets \mathit{r} \% |M|$
				\State \Return $t(n)$
			\EndIf
		\EndFunction
	\end{algorithmic}
\end{algorithm}

Note that we use the statement $\mathcal{K} \mathcal{M}_t$ to represent applying functor $\mathcal{K}$ to category $\mathcal{M}_t$ and get the result category. The detailed steps of the functor execution can be found before in the place of defining the stuff.

@@

A situation after some reinforcements may be the one demonstrated in Figure 2.

\begin{figure}[h!]
	\centering
	\begin{tikzpicture}[commutative diagrams/every diagram]
		\node (K0) at (90:2.3cm) {$\mathcal{K}_0$};
		\node (K1) at (90+72:2cm) {$\mathcal{K}_1$};
		\node (K2) at (90+2*72:2cm) {$\mathcal{K}_2$};
		\node (K3) at (90+3*72:2cm) {$\mathcal{K}_3$};
		\node (Kn) at (90+4*72:2cm) {$\mathcal{K}_N$};

@@

To summarize, we can illustrate the entire process of the workflow of the algorithm with the following diagram on the next page (see Figure 3).

\begin{figure}[h!]
	\centering
\begin{tikzpicture}[>=latex', font={\sf \small}]
	\def\smbwd{2cm}

	\node (start) at (0,0) [draw, terminal,
	minimum width = \smbwd,
	minimum height = 0.5cm] {Start};

	\node (parse) at (0,-1.5) [draw, process,
	minimum width = \smbwd,
	minimum height = 0.5cm] {\text{Get the representation of task from user}};

	\node (l-pool) at (-3,-3) [draw, storage,
	minimum width = \smbwd,
	minimum height = 0.5cm] {$\mathcal{M}_t$};

	\node (r-pool) at (3,-3) [draw, storage,
	minimum width = \smbwd,
	minimum height = 0.5cm] {$\mathcal{R}$};

	\node (rec-functor) at (-3,-4.5) [draw, predproc,
	minimum width = \smbwd,
	minimum height = 0.5cm] {$\mathcal{F}_t$};

	\node (random) at (-6,-4.5) [draw, process,
	minimum width = \smbwd,
	minimum height = 0.5cm] {$\mathit{r}$};

	\node (k-base) at (3,-4.5) [draw, storage,
	minimum width = \smbwd,
	minimum height = 0.5cm] {Knowledge base $\mathcal{Q}$};

	\node (success-val) at (0,-6) [draw, process,
	minimum width = \smbwd,
	minimum height = 0.5cm] {$s = {\exists n\in [0,\omega ), \mathcal{R}\subseteq \text{Ob}\left(\mathcal{M}_n\right)}$};

	\node (success-check) at (0,-7.5) [draw, decision,
	minimum width = \smbwd,
	minimum height = 0.5cm] {$s$};

	\node (reinforcement) at (3,-9) [draw, predproc,
	minimum width = \smbwd,
	minimum height = 0.5cm] {Reinforcement};

	\node (write-rh) at (-3,-9) [draw, process,
	minimum width = \smbwd,
	minimum height = 0.5cm] {Generate steps};

	\node (steps) at (-6,-9) [draw, storage,
	minimum width = \smbwd,
	minimum height = 0.5cm] {Proof steps};

	\node (forgetful-para) at (7,-9) [draw, process,
	minimum width = \smbwd,
	minimum height = 0.5cm] {$E(\mathit{t})$};

	\node (end) at (0,-10.5) [draw, terminal,
	minimum width = \smbwd,
	minimum height = 0.5cm] {End};

	\coordinate (point1) at (-7.5,-7.5);
	%% \coordinate (point2) at (-7,-3);

	\draw[->] (start) -- (parse);
	\draw[->] (parse) -| (l-pool);
	\draw[->] (parse) -| (r-pool);
	\draw[->] (random) -- (rec-functor);
	\draw[->] (l-pool) -- (rec-functor);
	\draw[->] (k-base) -- (rec-functor);
	\draw[->] (l-pool) -| (success-val);
	\draw[->] (r-pool) -| (success-val);
	\draw[->] (success-val) -- (success-check);
	%% \draw[->] (success-check) -| node[above]{No} (l-pool);
	\draw[->] (success-check) -- node[above]{No} (point1);
	\draw[->] (point1) |- (l-pool);
	\draw[->] (success-check) |- node[below]{Yes} (write-rh);
	\draw[->] (success-check) |- node[below]{Yes} (reinforcement);
	\draw[->] (write-rh) -- (steps);
	\draw[->] (write-rh) |- (end);
	\draw[->] (forgetful-para) -- (reinforcement);
	\draw[->] (reinforcement) -- (k-base);
	\draw[->] (reinforcement) |- (end);
\end{tikzpicture}
\caption{The flowchart of the entire algorithm.}
\end{figure}

@@

\section{Conclusions and Future Work}

@@

\section{Acknowledgments}
I express my sincerest gratitude to my tutor Prof. Shao Xinhui, the Yingcai Project of P.R.China, which enables senior high school students to conduct science research, my senior high academy Northeast Yucai School (NEYC) of P.R.China and the corresponding university Northern Eastern University (NEU) of P.R.China. I also thank my parents for encouraging me to conduct this research project.

Additionally, I had few knowledge about \LaTeX before writing this paper. The $lshort$ documentation and the \TeX Stack Exchange helped me a lot. Here I express my genuine thankfulness to the online resources and documentation of the used \LaTeX packages and software components.

\bibliographystyle{acm}
\bibliography{paper}
\end{document}
